---
title: 'EDS 223: assignment 3'
author: "Melissa Widas"
date: "2023-11-18"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
“In February 2021, the state of Texas suffered a major power crisis, which came about as a result of three severe winter storms sweeping across the United States on February 10–11, 13–17, and 15–20.”[^wiki] For more background, check out these [engineering](https://www.youtube.com/watch?v=08mwXICY4JM&ab_channel=PracticalEngineering) and [political](https://www.youtube.com/watch?v=Zcrsgdl_hP0&ab_channel=Vox) perspectives. 

For this assignment, you are tasked with:  
-   estimating the number of homes in Houston that lost power as a result of the first two storms  
-   investigating if socioeconomic factors are predictors of communities recovery from a power outage  

Your analysis will be based on remotely-sensed night lights data, acquired from the [Visible Infrared Imaging Radiometer Suite (VIIRS)](https://en.wikipedia.org/wiki/Visible_Infrared_Imaging_Radiometer_Suite) onboard the Suomi satellite. In particular, you will use the VNP46A1 to detect differences in night lights before and after the storm to identify areas that lost electric power.

To determine the number of homes that lost power, you link (spatially join) these areas with [OpenStreetMap](https://www.openstreetmap.org/#map=4/38.01/-95.84) data on buildings and roads. 

To investigate potential socioeconomic factors that influenced recovery, you will link your analysis with data from the US Census Bureau. 

##### Learning objectives:
-   load vector/raster data  
-   simple raster operations  
-   simple vector operations  
-   spatial joins  

### Data
#### Night lights
Use NASA's Worldview to explore the data around the day of the storm. There are several days with too much cloud cover to be useful, but 2021-02-07 and 2021-02-16 provide two clear, contrasting images to visualize the extent of the power outage in Texas.

VIIRS data is distributed through NASA's [Level-1 and Atmospheric Archive & Distribution System Distributed Active Archive Center (LAADS DAAC)](https://ladsweb.modaps.eosdis.nasa.gov/). Many NASA Earth data products are distributed in 10x10 degree tiles in sinusoidal equal-area projection. Tiles are identified by their horizontal and vertical position in the grid. Houston lies on the border of tiles h08v05 and h08v06. We therefore need to download two tiles per date.

As you're learning in EDS 220, accessing, downloading, and preparing remote sensing data is a skill in it's own right! To prevent this assignment from being a large data wrangling challenge, we have downloaded and prepped the following files for you to work with, stored in the `VNP46A1` folder.\

-   `VNP46A1.A2021038.h08v05.001.2021039064328.h5.tif`: tile h08v05, collected on 2021-02-07  
-   `VNP46A1.A2021038.h08v06.001.2021039064329.h5.tif`: tile h08v06, collected on 2021-02-07  
-   `VNP46A1.A2021047.h08v05.001.2021048091106.h5.tif`: tile h08v05, collected on 2021-02-16  
-   `VNP46A1.A2021047.h08v06.001.2021048091105.h5.tif`: tile h08v06, collected on 2021-02-16  

#### Roads
Typically highways account for a large portion of the night lights observable from space (see Google's [Earth at Night](https://earth.google.com/web/@27.44405464,-84.7693044,206.63660162a,8916361.52264659d,35y,0h,0t,0r/data=CiQSIhIgMGY3ZTJkYzdlOGExMTFlNjk5MGQ2ZjgxOGQ2OWE2ZTc)). To minimize falsely identifying areas with reduced traffic as areas without power, we will ignore areas near highways.

[OpenStreetMap (OSM)](https://planet.openstreetmap.org/) is a collaborative project which creates publicly available geographic data of the world. Ingesting this data into a database where it can be subsetted and processed is a large undertaking. Fortunately, third party companies redistribute OSM data. We used [Geofabrik's download sites](https://download.geofabrik.de/) to retrieve a shapefile of all highways in Texas and prepared a Geopackage (`.gpkg` file) containing just the subset of roads that intersect the Houston metropolitan area.\  

- `gis_osm_roads_free_1.gpkg`

[^wiki]:Wikipedia. 2021. “2021 Texas power crisis.” Last modified October 2, 2021. https://en.wikipedia.org/wiki/2021_Texas_power_crisis.

#### Houses
We can also obtain building data from OpenStreetMap. We again downloaded from Geofabrick and prepared a GeoPackage containing only houses in the Houston metropolitan area.\

- `gis_osm_buildings_a_free_1.gpkg`

#### Socioeconomic
We cannot readily get socioeconomic information for every home, so instead we obtained data from the [U.S. Census Bureau's American Community Survey](https://www.census.gov/programs-surveys/acs) for census tracts in 2019. The *folder* `ACS_2019_5YR_TRACT_48.gdb` is an ArcGIS ["file geodatabase"](https://desktop.arcgis.com/en/arcmap/latest/manage-data/administer-file-gdbs/file-geodatabases.htm), a multi-file proprietary format that's roughly analogous to a GeoPackage file.\

You can use `st_layers()` to explore the contents of the geodatabase. Each layer contains a subset of the fields documents in the [ACS metadata](https://www2.census.gov/geo/docs/maps-data/data/tiger/prejoined/ACSMetadata2011.txt).\

The geodatabase contains a layer holding the geometry information, separate from the layers holding the ACS attributes. You have to combine the geometry with the attributes to get a feature layer that `sf` can use. 

## Assignment

Below is an outline of the steps you should consider taking to achieve the assignment tasks. 

#### Find locations of blackouts
For improved computational efficiency and easier interoperability with `sf`, I recommend using the `stars` package for raster handling.\

```{r message= FALSE, warning= FALSE}
library(tidyverse)
library(sf)
library(stars)
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
library(raster)
library(rgdal)
library(ggspatial)
library(cowplot)
```


##### combine the data (5 points)
-   read in night lights tiles\
-    combine tiles into a single `stars` object for each date (2021-02-07 and 2021-02-16)\
    -   hint: use `st_mosaic`\
    
```{r include=TRUE}
# read in night lights tiles
night1 <- read_stars('data/VNP46A1/VNP46A1.A2021038.h08v05.001.2021039064328.tif')
night2 <- read_stars('data/VNP46A1/VNP46A1.A2021038.h08v06.001.2021039064329.tif')
night3 <- read_stars('data/VNP46A1/VNP46A1.A2021047.h08v05.001.2021048091106.tif')
night4 <- read_stars('data/VNP46A1/VNP46A1.A2021047.h08v06.001.2021048091105.tif')

# check class of night light tiles
class(night1)

# combine tiles into stars object for 2021-02-07 and 2021-02-16
night_07 <- st_mosaic(night1, night2)
night_16 <- st_mosaic(night3, night4)

# view 2021-02-07 raster
plot(night_07, main = "Raster of 2021-02-07")
```
    
##### create a blackout mask (10 points)
-   find the change in night lights intensity (presumably) caused by the storm\
-   reclassify the difference raster, assuming that any location that experienced a drop of more than 200 nW cm^-2^sr^-1^ experienced a blackout\
-   assign `NA` to all locations that experienced a drop of *less* than 200 nW cm^-2^sr^-1^\

```{r include=TRUE}
# find the change in night light intensity between the two dates 
light_difference <- night_07-night_16

# check class of difference data created 
class(light_difference)

# reclassifying raster to show blackouts above 200, and label areas that experienced an outage outage, and that didn't NA
light_difference_outage <- cut(light_difference, c(200, Inf), labels = 'outage')

#show where blackouts were experienced
plot(light_difference_outage, main = "Areas that Experienced a Blackout")

```


##### vectorize the mask (5 points)
-   use `st_as_sf()` to vectorize the blackout mask\
-   fix any invalid geometries using `st_make_valid`

```{r include=TRUE}
# vectorize the blackout mask to make this a spatial feature
blackouts <- st_as_sf(light_difference_outage) %>% 
  st_make_valid()

# represents outages in houston above 200
plot(blackouts, main = "Houston, TX Metropolitan Areas that Experienced a Blackout")
```


##### crop the vectorized map to our region of interest (10 points)
-   define the Houston metropolitan area with the following coordinates\
    -   (-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29)\
-   turn these coordinates into a polygon using `st_polygon`\
-   convert the polygon into a simple feature collection using `st_sfc()` and assign a CRS\
    -   hint: because we are using this polygon to crop the night lights data it needs the same CRS\
-   crop (spatially subset) the blackout mask to our region of interest\ 
-   re-project the cropped blackout dataset to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area)\

```{r include=TRUE}
# defining houston coordinates
houston <- cbind(x = c(-96.5, -96.5, -94.5, -94.5, -96.5), 
                         y = c(29, 30.5, 30.5, 29, 29))

#turn houston coordinates into polygon
houston <- st_sfc(st_polygon(list(houston)), crs = 4326)

# create a mask for the houston area 
houston_mask <- st_intersects(blackouts, houston, sparse = FALSE)

# crop the blackout_mask to the houston region of interest 
houston_blackouts <- blackouts[houston_mask,]

# reprojecting houston blackouts to Texas specific Datum
houston_blackouts_reproj <- st_transform(houston_blackouts, crs = 3083)

# plot houston blackouts 
plot(houston_blackouts_reproj, main = "Blackouts in Houston")
```


##### exclude highways from blackout mask (10 points)
The roads geopackage includes data on roads other than highways. However, we can avoid reading in data we don't need by taking advantage of `st_read`'s ability to subset using a SQL query.\

-   define SQL query\
-   load just highway data from geopackage using `st_read`\
  -   reproject data to EPSG:3083\
-   identify areas within 200m of all highways using `st_buffer`\
    -   hint: `st_buffer` produces undissolved buffers, use `st_union` to dissolve them\
-   find areas that experienced blackouts that are further than 200m from a highway

`query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"`\
`highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = query)`

```{r include=TRUE}
# creates the query to select motorways
highway_query <- "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'"

# load highway data that fulfills query 
highways <- st_read("data/gis_osm_roads_free_1.gpkg", query = highway_query)

# reproject highwyas to crs 3083
highways_reproj <- st_transform(highways, crs = 3083)


# identify areas within 200m of all highways
highways_buffer <- st_buffer(highways_reproj, dist = 200) %>% 
  st_union()

# find areas that had blackouts that are further than 200m 
blackouts_outside_highway <- st_difference(houston_blackouts_reproj, highways_buffer)

# blackouts outside of the highway buffer should be less than all of the houston blackouts
nrow(blackouts_outside_highway)<nrow(houston_blackouts_reproj)
```


#### Find homes impacted by blackouts

##### load buildings data (10 points)
-   load buildings dataset using `st_read` and the following SQL query to select only residential buildings\
  -   hint: reproject data to EPSG:3083\

`SELECT *`\ 
`FROM gis_osm_buildings_a_free_1`\
`WHERE (type IS NULL AND name IS NULL)`\
`OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')`\

```{r include=TRUE}
# creates the query to select residential buildings
residential_query <- "SELECT * FROM gis_osm_buildings_a_free_1 WHERE (type IS NULL AND name IS NULL) OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')"

# load buildings data that fulfills query 
residential <- st_read("data/gis_osm_buildings_a_free_1.gpkg", query = residential_query)

# reproject buildings to crs 3083
residential_reproj <- st_transform(residential, crs = 3083)
```

##### find homes in blackout areas (20 points)
-   filter to homes within blackout areas\
-   count number of impacted homes\

```{r include=TRUE}
# dataframe where residential buildings are kept that are in houston outside of our highway buffer
blackout_residential <- residential_reproj[blackouts_outside_highway, , op = st_intersects]

# count number of buildings in blackout areas 
nrow(blackout_residential)

#counting the number of residential buildings that got hit
print(paste0(nrow(blackout_residential), ' residential buildings were affected by the blackouts'))

```
#### Investigate socioeconomic factors

##### load ACS data (10 points)
-   use `st_read()` to load the geodatabase layers\
-   geometries are stored in the `ACS_2019_5YR_TRACT_48_TEXAS` layer\
-   income data is stored in the `X19_INCOME` layer\
-   select the median income field `B19013e1`\
  -   hint: reproject data to EPSG:3083\

```{r include=TRUE}
# load data that contains census tract geometries
texas_census <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", layer =  "ACS_2019_5YR_TRACT_48_TEXAS")

# reproject census data to crs 3083
texas_census_reproj <- st_transform(texas_census, crs = 3083)

# load data that contains census tract data and income data
texas_income <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb", layer =  "X19_INCOME")

# create data frame that contains median income data and geoID
texas_median_income <- texas_income %>% 
  dplyr::select(B19013e1, GEOID) %>% 
  rename(median_income = B19013e1, GEOID_Data = GEOID)
```


##### determine which census tracts experienced blackouts (10 points)
-   join the income data to the census tract geometries\
  -   hint: make sure to join by geometry ID\
-   spatially join census tract data with buildings determined to be impacted by blackouts\
-   find which census tracts had blackouts\

```{r include=TRUE}
# join the income data to census tract geometries in Texas
texas_income_census <- left_join(texas_census_reproj, texas_median_income, by = "GEOID_Data")

# join the census tract data with residential buildings determined to be impacted by blackouts
texas_census_blackouts <- st_filter(texas_income_census, blackout_residential) 
```


##### compare incomes of impacted tracts to unimpacted tracts (10 points)
-   create a map of median income by census tract, designating which tracts had blackouts\
-   plot the distribution of income in impacted and unimpacted tracts
-   write approx. 100 words summarizing your results and discussing any limitations to this study


```{r}
print(paste("Income of Census tracts in Texas is the same CRS as Houston boundary:", (st_crs(houston)==st_crs(texas_income_census))))

#transforming houston boundary to Texas datum
houston_reproj <- st_transform(houston, crs = 3083)

# data frame that has all median income is in houston census tracts
houston_income_census <- st_filter(texas_income_census, houston_reproj)

# create visualization
median_income_map <- ggplot() +
  geom_sf(data = houston_income_census, # plot the median income of all of houston
          aes(fill = median_income)) +
  scale_fill_viridis_c() +
  geom_sf(data = texas_census_blackouts, aes(color = "red", # outline census tracts that experienced blackouts in red
                                       fill = median_income)) +
  scale_color_hue(labels = c("Blackout Census Tracts")) + # alter legend titles
  labs(fill='Median Income') +
  theme_void() +
  annotation_north_arrow( # add north arrow
  height = unit(1.5, "cm"),
  width = unit(1.5, "cm"),
  pad_x = unit(7.5, "cm"),
  pad_y = unit(0.15, "cm"),
  rotation = NULL,
  style = north_arrow_fancy_orienteering
) +
  annotation_scale( # add scale bar
  plot_unit = NULL,
  bar_cols = c("black", "white"),
  line_width = 1,
  height = unit(0.15, "cm"),
  pad_x = unit(2, "cm"),
  pad_y = unit(0.15, "cm"),
  text_pad = unit(0.15, "cm"),
  text_cex = 0.7,
  text_face = NULL,
  text_family = "",
  tick_height = 0.6
) +
  labs(title= "Median Income of Census Tracts in Houston, TX", color = "") # add map title
  
median_income_map
```




```{r}

# find areas that did not have blackouts within the houston area 
unimpacted <- houston_income_census[texas_census_blackouts, , op = st_difference]

# create a histogram showing median income spread for homes in houston not impacted by blackouts 
unimpacted_figure <- ggplot(data = unimpacted, aes(median_income)) +
  geom_histogram(bins=50, fill = '#481467FF', color = 'white') +
  labs(title = "Unimpacted by Blackouts", x = "Median Income ($)", y = "Census Tracts") +
  theme_classic() +
  ylim(0,105)
  
#unimpacted_figure

# create a histogram showing median income spread for homes in houston impacted by blackouts
impacted_figure <- ggplot(data = texas_census_blackouts, aes(median_income)) +
  geom_histogram(bins=50, fill = '#481467FF', color = 'white') +
  labs(title = "Impacted by Blackouts", x = "Median Income ($)", y = "Census Tracts") +
  theme_classic() +
  ylim(0,105)

# position histograms side by side for easier comparison
impact_unimpact_plot <- plot_grid(unimpacted_figure, impacted_figure, labels = c('', ''), label_size = 12)

impact_unimpact_plot
```
**The median income of homes in Houston, TX affected and not affected by blackouts does not appear to have a significant difference. The blackouts do appear to have centered around the center of the city, but this area also has a wide variety of median incomes. Additionally, when comparing the median incomes of homes affected and not affected the distributions appear to be similar for the spread of wealth. Residents not impacted by the storm do appear to have slightly higher median incomes which could coincide with living farther outside the city center in the suburbs of Houston. A limitation of this analysis is that we did reduce the number of buildings counted that fell close to highways, which could be areas that have lower median incomes. Additionally, in exploring the power outages in Houston, TX we only compared the blackouts from the first two storms, without accounting for the third storm. Without accounting for the third storm we would be underestimating the total number of outages that occured.**
